---
title: "Graphical Markov Models with Mixed Graphs in R"
author: "Kayvan Sadeghi and Giovanni M. Marchetti"
institute: "DISIA, University of Florence"
date: "last-modified"
lang: en
bibliography: references.bib
format: 
  html: default
#  pdf: default
metadata-files: 
  - style.yml
---
# Introduction

We provide a short tutorial illustrating the functions in the package **ggm** that deal with **mixed graphs**.
A mixed graph can have 3 types of edges and they are 
important because they capture the modified
independence structure after marginalization over and conditioning on nodes of directed acyclic graphs.

We provide functions 

- to verify whether a mixed graph implies that $A$ is independent of $B$ given $C$ for any disjoint sets of nodes, and 
- to generate maximal graphs inducing the same independence structure of non-maximal graphs. 
- Finally, we provide functions to decide on the Markov equivalence of two graphs with the same node set but different type of edges.


# Introduction and background

**Graphical Markov models** have become a part of the mainstream of statistical theory and application in recent years. These models use graphs to represent conditional independencies among sets of random variables. Nodes of the graph correspond to random variables and edges to some type of conditional dependencies.

## Directed acyclic graphs

In the literature of graphical models the two most used classes of graphs  are the directed
acyclic graphs (DAGs) and the undirected graphs. DAGs have been proven useful, among other things,
to specify the data generating processes when the variables satisfy an underlying partial ordering.

 For instance, suppose that we have 4  observed variables $Y$, the ratio of systolic to diastolic blood pressure, $X$ the diastolic blood pressure, both on log scale,
  $Z$, the body mass and $W$, the age, and that a possible generating process is the following  linear recursive regression model
$$
\begin{split}
Y &=& \gamma_{YZ} Z + \gamma_{YU} U + \epsilon_Y\\
X &=& \gamma_{XW} W + \gamma_{XU} U + \epsilon_X\\
Z &=& \gamma_{ZV} W + \epsilon_Z\\
W &=& \epsilon_W; \; U = \epsilon_U;
\end{split}
$$
where all the variables are mean-centered and the $\epsilon$s are zero mean, mutually independent Gaussian random errors. In this model we
assume that there exist a genetic factor $U$ influencing
the ratio and levels of blood pressure.

Then, this model can be represented by the DAG in the following Figure, with nodes associated with the variables
and edges, indicating the dependencies, represented  by
the regression coefficients $\gamma$s.

```{r, warnings = FALSE, echo=FALSE, message=FALSE, fig.width = 2, fig.height= 2}
require(ggm)
G1 = DAG( Y ~ U + Z, X ~ U+W, Z~ W)
drawGraph(G1, coor = matrix(c(10,90, 50,50,90,90,10,10, 90,10), 
                            byrow = TRUE, ncol = 2))
```

From the graph it is seen for instance
that the ratio of the two blood pressures
is directly influenced by body mass but not by age. Thus a
consequence of the model is that the variables must satisfy
a set of conditional independencies: for example the ratio of the blood pressure $Y$
is independent of age $W$ given body mass $Z$, written as $Y \ci W \mid Z$.

 A remarkable result is
that the independencies can be deduced from the graph alone, without reference to the equations, by using a criterion called **d-separation**. 

In fact, in the graph of previous Figure the nodes $Y$ and $W$ are d-separated given $Z$ as can be checked
using special graph algorithms included for example in packages **gRain** and **ggm**.

For more details on   DAG models and their implementation in R see the extensive discussion in \citet{hoi12}.

## Hidden variables and induced graphs

The model has  4 observed variables but includes an unobserved variable,  that is the genetic factor $U$. When  $U$ is hidden the model becomes
  \begin{eqnarray*}
Y &=& \gamma_{YZ} Z + \eta_Y \\
X &=& \gamma_{XW} W + \eta_X \\
Z &=& \gamma_{ZV} W + \epsilon_Z\\
W &=& \epsilon_W;
\end{eqnarray*}
with two correlated errors $\eta_{Y} =  \gamma_{YU} U + \epsilon_Y$ and $\eta_X =  \gamma_{XU} U + \epsilon_X$,
such that $\mathrm{cov}(\eta_Y, \eta_X) = \omega){YX}$.

As a consequence the model is still a recursive model and the parameters have a regression parameter interpretation, but contains some correlated residuals.

The induced model is said to be obtained
after marginalization over $U$.  In this model some of the original independencies are lost,
but we can observe the implied independencies
$Y\ci W|Z$ and $X \ci Z \mid  W$. Also it can be shown that
in a DAG model defined for the 4 observed variables
is impossible to represent such independencies. Therefore, we say
that  DAG models are not **stable under marginalization**.

A mixed graph with arrows and arcs, shown in the following Figure can however be used to represent the induced independence  model after marginalization over $U$.

```{r, fig.width = 2, fig.height = 2, echo =FALSE}
G2 = makeMG(dg = DAG(Y ~ Z, X ~ W, Z ~ W), bg = UG(~ Y*X))
drawGraph(G2, coor = matrix(c(10,90, 90,90,10,10, 90,10), 
                            byrow = TRUE, ncol = 2))
```

In this representation, beside the arrows, represented by the
$\gamma$s, we have the arc $Y \leftrightarrow X$ associated with the (partial) correlation $\omega_{YX}$.

The resulting graph belongs to a class of models
called **regression chain graph models**. This class generalizes the recursive generating
process of DAGs by permitting joint responses, coupled in the graph by arcs, and thus appears an essential extension for applications.

The class of regression graphs is however itself 
not stable under marginalization.
For instance,  suppose that the  generating process for the blood pressure data is defined by the more  general regression chain graph 
```{r, fig.width = 2, fig.height = 2, echo =FALSE}
G3 = makeMG(dg = DAG(Y ~ Z, X ~ W, Z ~ W, X~L, Z~L), bg = UG(~ Y*X))
drawGraph(G3, coor = matrix(c(10,90, 90,90,10,10, 90,10, 50, 50), 
                            byrow = TRUE, ncol = 2))
```

where $L$ is a further variable representing a common hidden cause of systolic good pressure and body mass.

Then, after marginalization over $L$,  the model can still be described  by a a linear system of equations with correlated residuals and can be represented by the mixed graph below 

```{r, fig.width = 2, fig.height = 2, echo =FALSE}
G4 = makeMG(dg = DAG(Y ~ Z, X ~ W, Z ~ W), bg = UG(~ Y*X + Z*X))
drawGraph(G4, coor = matrix(c(10,90, 90,90,10,10, 90,10), 
                            byrow = TRUE, ncol = 2))
```

but the resulting graph is not a DAG, nor a regression graph because
it contains the pair of variables $(Y, Z)$  coupled both by a directed edge and by a path composed by bi-directed arcs. 

Thus $Y$ cannot be interpreted as a pure response to $Z$ and in addition $Y$ and $Z$ are not two joint responses.


## Stable mixed graphs

The previous illustrations show that when there are unobserved
variables, DAG or regression chain graph models are no longer appropriate. The discussion could be extended
to situations where there are some selection variables, that is hidden variables that are conditioned on.

This motivates the introduction of a more general class of \emph{mixed graphs}, which contains
three types of edge, denoted by lines, $-\!\!-$, arrows, $\rightarrow$, and arcs (bi-directed arrows), $\leftrightarrow$.

There are at least three known classes of mixed graphs without self loops that remain in the same class, i.e.
that are **stable under marginalisation and conditioning**. 

- The largest one is that of **ribbonless graphs** (RGs),
 \citep{sad112}, defined as a modification of MC-graphs \citep{kos02}.
 
- Then, there is the subclass of **summary graphs** (SGs) \citep{wer11},

- and finally the smallest class of the **ancestral graphs** (AGs),  \citep{ric02}.


## Four tasks of the current paper

In this paper, we focus on the implementation of four important tasks performed on the class of mixed graphs in R:

- Generating different types of stable mixed graphs after marginalisation and conditioning.
- Verifying whether an independency of form $Y\ci W\mid Z$ holds by using a separation criterion called $m$-separation.
- Generating a graph that induces the same independence model as an input mixed graph such that the generated graph is **maximal**, i.e.\ each missing edge of the generated graph implies at least an independence statement.
- Verifying whether two graphs are **Markov equivalent**, i.e.\ they induce the same independencies, and whether there is a Markov equivalent graph of a different type to an input graph of a specific type.

In the next section we give the details on how general mixed graphs are defined in package `ggm`. Next in each of the following sections we respectively deal with each of the tasks described above. For each task we give a brief introduction at the beginning of its corresponding section.

## Package `ggm`

The previous tasks are illustrated by using a set of new functions introduced into the `R` package `ggm`. Some of the functions
generalize previous contributions of `ggm` discussed in \citet{mar06}.

Package `ggm` has been improved and it is
now more integrated with other contributed packages related to graph theory, such as `graph`, `igraph` \citep{csa09}, and `gRbase` \citep{DetHo05}, which are now required for representing and plotting graphs.

Specifically, in addition to adjacency matrices, all the functions in the package now accept `graphNEL` and `igraph` objects as input, as well as a new character string representation. A more detailed list of available packages for graphical models
 can be found at CRAN Task View [gRaphical Models in R](http://cran.r-project.org/web/views/gR.html).


# Defining mixed graphs in R

For a comprehensive discussion on the ways of defining a directed acyclic a graph, see \citep{hoi12}.  A **mixed graph** is a more general graph type with at most three types of edge: directed, undirected and bi-directed, with possibly
multiple edges of different types connecting two nodes.  

In `ggm` we provide some special tools for mixed graphs that are not present in other packages. Here we briefly illustrate
some methods to define mixed graphs and we plot them with a new function, `plotGraph`, which uses a Tk GUI for basic interactive graph manipulation.

- The first method is based on a generalization of the adjacency matrix. 
- The second uses a descriptive vector and is easy to use for small graphs. 

- The third uses a special function `makeMG`
that allows to combine the directed, undirected and bi-directed components of a mixed graph.

## Adjacency matrices for mixed graphs

In the adjacency matrix of a mixed graph we code the three different edges with a binary indicator: 

- 1 for directed, 
- 10 for undirected and 
- 100 for bi-directed edges. 

When there are multiple edges the codes are **added**.

Thus the **adjacency matrix of a mixed graph**  $H$ with node set $N$ and edge set $F$ is an $|N|\times|N|$ matrix obtained as $A=B+S+W$ by adding three matrices $B = (b_{ij})$, $S = (s_{ij})$ and
$W = (w_{ij})$ defined by

$$
\begin{split}
b_{ij} &=&
  \begin{cases}
    1, & \text{ if and only if } i \rightarrow j \text{ in } H; \\
    0, & \text{otherwise.}\\
  \end{cases}\\
s_{ij}=s_{ji}&=&
  \begin{cases}
    10, & \text{if and only if } i -\!\!\!- j \text{ in } H; \\
    0, & \text{otherwise.}
  \end{cases}\\
w_{ij}=w_{ji}&=&
  \begin{cases}
    100, & \text{if and only if } i\leftrightarrow j \text{ in } H; \\
    0, & \text{otherwise.}
  \end{cases}
\end{split}
$$
Notice that because of the symmetric nature of lines and arcs $S$ and $W$ are symmetric, whereas $B$ is not necessarily symmetric.

For instance the following graph is a mixed graph (Notice that this mixed graph is not even ribbonless and hence of not much interest, and has been only introduced for presentation purposes):
```{r, warning= FALSE}
mg <- matrix(c( 0, 101,   0,   0, 110,
              100,   0, 100,   0,   1,
                0, 110,   0,   1,   0,
                0,   0,   1,   0, 100,
              110,   0,   0, 100,   0), 5,5, byrow = TRUE)
N <- c("X","Y","Z","W","Q")
dimnames(mg) <- list(N, N)
drawGraph(mg, coor = matrix(c(10, 50, 30,90, 90,90,  
                              90, 10, 30, 10),
                            ncol = 2, byrow  = TRUE))
```
The graph can be plotted also with `plotGraph(mg)`.

\paragraph{Defining mixed graphs by using vectors.} A more convenient way of defining small mixed graphs is based on a simple  vector coding as follows. The graph is defined by a character vector of length $3f$, where
$f = |F|$ is the number of edges, and the vector contains a sequence of triples $\langle$\code{type, label1, label2}$\rangle$,
where the \code{type} is the edge type and \code{label1} and \code{label2} are the labels of the two nodes.
The edge type accepts \code{"a"} for an directed arrow ,  \code{"b"} for an arcs and \code{"l"} for a line.
Notice that isolated nodes may not be created by this method. For example, the vector representation of the previous mixed graph is
<<>>=
mgv <- c("b","X","Y","a","X","Y","l","X","Q",
         "b","Q","X","a","Y","Q","b","Y","Z",
         "a","Z","W","a","W","Z","b","W","Q")
@
Once again as in the DAG case we can use \code{plotGraph(mgv)}
to plot the defined graph.
\paragraph{Mixed graph using the function \code{makeMG}.} Finally  the
adjacency matrix of a mixed graph may be built up with the function \code{makeMG}. This function requires
three arguments \code{dg}, \code{ug} and \code{bg}, corresponding respectively
to the three adjacency matrices $B$, $S$ and $W$
composing the mixed graph. These may also be obtained by the constructor functions
\code{DG} and \code{UG} of \pkg{ggm} for directed and undirected graphs respectively. Thus for the
previous mixed graph we can issue the command
<<>>=
mg <- makeMG(dg = DG(Y~X, Z~W, W~Z),
            ug = UG(~ X*Q),
            bg = UG(~ Y*X + X*Q + Q*W + Y*Z))
@
obtaining the same adjacency matrix (up to a permutation).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Generating stable mixed graphs}
There are four general classes of stable mixed graphs.

Ribbonless graphs are mixed graphs without a specific set of subgraphs called ribbons. The lack of ribbons ensures that, for any RG, there is a DAG whose independence structure, i.e.\ the set of all conditional independence statements that it induces, after marginalisation over and conditioning on two disjoint subsets of its node set can be represented by the given RG. This is particularly essential as it shows that the independence structures corresponding to RGs are probabilistic, that is there exists a probability distribution $P$ that is faithful with respect to any RG, i.e.\ for random vectors $X_A$, $X_B$, and $X_C$ with probability distribution $P$, $X_A\cip X_B\cd X_C$ if and only if $\langle A,B\cd C\rangle$ is in the induced independence structure by the graph. This probability distribution is the marginal and conditional of a probability distribution that is faithful to the generating DAG.

Summary graphs have neither arrowheads pointing to lines (i.e.\ $\arc\circ\ful$ or $\fra\circ\ful$) nor directed cycles with all its arrows pointing towards one direction. Ancestral graphs have these two constraint together with an additional constraint that these graphs do not contain so-called \emph{bows}, see \citet{ric02}. 

However, notice that for some ribbonless and summary graphs the corresponding parametrisation is sometimes not available even in the case of standard joint Gaussian distribution.

If we suppose that stable mixed graphs are only used to represent the independence structure after marginalisation and conditioning, we can consider all types as equally appropriate. However, each of the three types has been used in different context and for different purposes. RGs have been introduced in order to straightforwardly deal with the problem of finding a class of graphs that is closed under marginalisation and conditioning by a simple process of deriving them from DAGs. SGs are used when the generating DAG is known and to trace the effects in the sets of regressions as described above. AGs are simple graphs, meaning that they do not contain multiple edges and the lack of bows ensures that they satisfy many desirable statistical properties. 

In addition, when one traces the effects in regression models with latent and selection variables (as described in the introduction) ribbonless graphs are more alerting to possible distortions (due to indirect effects) than summary graphs, and summary graphs are more alerting than ancestral graphs; see also \citet{werc08}. For the exact definition and a thorough discussion of all such graphs, see \citet{sad112}.


\citet{sad112} also defines the algorithms generating stable mixed graphs of a specific type for a given DAG or for a stable mixed graphs of the same type after marginalisation and conditioning such that they induce the marginal and conditional DAG-independence structure. We implement these algorithms in this paper.

%\section{Defining directed acyclic graphs}
%We start by illustrating some ways of defining directed acyclic graphs in R.
%We shall use four graph representations, that is
%(a) by \code{graphNEL} objects, (b) by \code{igraph} objects, (c)  by adjacency matrices, (d) by character
%vectors. These same representations will be extended to general mixed graphs in the next section.
%\paragraph{Defining DAGs as graphNEL objects.} The \pkg{graph} package provides the \code{graphNEL} class for
%graphs that are represented in terms of nodes and an edge list.
%These graphs can be either directed or undirected, with no multiple edges or self-loops. The \pkg{gRbase} package provides the function  \code{dag} for easily defining DAGs as
%\code{graphNEL} objects. One way in which the graph can be specified is by a list
%of formulas. For example, one can define a DAG by the following formula, where, for example, \code{~h*p*r} implies that $h$ has parents $p$ and $r$. For reference, we plot the DAG after the explanations related to the formula, by using the
%\pkg{ggm} function \code{plotGraph}.
%

%
%<<>>=% 2
%exdag <- dag(~r, ~q*r, ~p, ~h*p*r, ~l,
%             ~k*l, ~j*k, ~i*h*l*j)
%exdag
%@
%Notice that one cannot use integers as node labels in \code{graphNEL} objects.
%In addition, instead of ``*'' a ``:'' can be used in the specification. To retrieve the nodes and edges of the DAG the functions \code{nodes} and \code{edges} are used. Thus for instance,
%<<>>=% 3
%nodes(exdag)
%@
%To plot a DAG as a \code{graphNEL} object one can use the function \code{plot} in the \pkg{Rgraphviz} package or
%various plotting options within the \pkg{igraph} package (see below).
%Our recommended way is to use  a defined function \code{plotGraph} as follows:
%<<eval=false>>=
%plotGraph(exdag)
%@
%\begin{center}
%\includegraphics[scale = 0.45]{fig1}
%\end{center}
%\paragraph{Defining DAGs as igraph objects.} In package \pkg{igraph} there are many
%ways of defining DAGs as \code{igraph} objects; for the full list  see \citet{csa09}.
%Probably the most handy way of generating small graphs is to use \code{graph.formula}. In this function we use ``-'' as edges and ``+'' as arrowheads. For example, for the same DAG as before, we use
%<<>>= % 4
%exdag <- graph.formula(q+-r-+h+-p,
%         h-+i+-l-+k,   i+-j+-k)
%exdag
%@
%It is always possible to convert \code{igraph} graphs to \code{graphNEL} objects or back using the \code{igraph.to.graphNEL} and \code{igraph.from.graphNel} functions.
%
%Also, the package \pkg{igraph} contains several options for plotting graphs such as the function
% \code{tkplot}. This is a quite flexible function providing an interactive graph drawing facility based on Tcl-Tk.
%Our \code{plotGraph} function is just a convenient wrapper of this function.
%\paragraph{Defining DAGs with adjacency matrices.} Any graph can be defined by an \emph{adjacency matrix}.
% The adjacency matrix of a loopless directed graph $G$ with node set $V$ and edge set $E$  is the matrix $A=(a_{ij})$ of size $|V|\times |V|$ with zeros on the diagonal and
%$$
%a_{ij}=\begin{cases}
%       1, & \text{ if } i\fra j; \\
%       0  & \text{otherwise.}\\
%   \end{cases}
%$$
%An adjacency matrix can be defined directly without defining a \code{graphNEL} or  \code{igraph} object.
%For instance the adjacency matrix of the previous DAG can be created via:
%
%<<>>= % 5
%A <- matrix(0,8,8)
%A[8,7] <- A[8,5] <- A[6,5] <- A[5,1] <-
%A[4,1] <- A[4,3] <- A[2,1] <- A[3,2] <- 1
%A
%@
%All functions in package \pkg{ggm} accept adjacency matrices as defining graphs. If the node names are
%not provided, the functions would specify by default as labels the  integers $\langle 1,\dots,n\rangle$,
%where $n$ is the number nodes. Thus the defined adjacency matrix \code{A} defines the following graph:
%<<eval=FALSE>>= % 6
%plotGraph(A)
%@
%\begin{center}
%\includegraphics[scale = 0.45]{fig51n}
%\end{center}
%The node labels are obtained by including appropriate  \code{dimnames} to the matrix:
%<<>>=
%V <- c("i","j","k","l","h","p","q","r")
%dimnames(A) = list(V, V)
%@
%An adjacency matrix for the DAG can be also obtained using the function \code{DAG} in \pkg{ggm}.
%This function accepts as input a model formula an returns the adjacency matrix with the appropriate labels.
%For the same DAG, which we call \code{exdag}, we can write
%<<>>=
%exdag <- DAG(i ~ j+l+h, k~l, h~r+p, q ~ r)
%@
%The function also verifies whether the resulting graph is acyclic.
By ``generating graphs'' we mean applying the defined algorithms, e.g.\ those for generating stable mixed to graphs, in order to generate new graphs.
\paragraph{Functions to generate the three main types of stable mixed graphs.}
Three main functions \code{RG}, \code{SG}, and \code{AG} are available to generate and plot ribbonless, summary, and ancestral graphs from DAGs,  using the algorithms in \citet{sad112}. These algorithms look for the paths with three nodes and
two edges in the graph whose inner nodes are being marginalized over or condition on, and generate appropriate edges between the endpoints. These have two important properties: (a) they are well-defined in the sense that
the process can be performed in any order and produces always the same final graph; (b) the generated graphs
induce the modified independence structure after marginalization and conditioning; see \citet{sad112} for more details.

The functions  \code{RG}, \code{SG}, and \code{AG} all have three arguments: \code{a} the given input graph,
\code{M}, the marginalization set and \code{C}, the conditioning set. The graph may be of class \code{"graphNEL"} or  of class \code{"igraph"} or may be represented by a character vector, or by an adjacency matrix, as explained in the previous sections. The sets \code{M} and \code {C} (default \code{c()}) must be disjoint vectors of node labels, and
they may possibly be empty sets.
The output is always the adjacency matrix of the generated graph. There are two additional
logical arguments \code{showmat} and \code{plot} to specify whether the adjacency matrix must be explicitly printed
(default \code{TRUE}) and  the graph must be plotted (default \code{FALSE}).

\paragraph{Some examples.} We start from a DAG defined in two ways, as an adjacency matrix and as a
character vector:
<<>>= % 9
ex <- matrix(c(0,1,0,0,0,0,0,0,
               0,0,1,0,0,0,0,0,
               0,0,0,0,0,0,0,0,
               0,0,1,0,1,0,1,0,
               0,0,0,0,0,1,0,0,
               0,0,0,0,0,0,0,0,
               0,0,0,0,0,1,0,0,
               0,0,0,0,0,1,1,0),8,8,byrow=TRUE)
exvec <- c("a",1,2,"a",2,3,"a",4,3,
           "a",4,5,"a",4,7,"a",5,6,
           "a",7,6,"a",8,6,"a",8,7)
@
<<eval=FALSE>>= % 10
plotGraph(ex)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setkeys{Gin}{width=0.5\textwidth}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\includegraphics[scale = 0.45]{fig2}
\end{center}
Then we define two disjoint sets $M$ and $C$ to marginalize over and condition on
<<>>=
M <- c(5,8)
C <- 3
@
and we generate the ribbonless, summary and ancestral graphs from the DAG with the associated plot.
<<eval = false>>= % 11
RG(ex, M, C, plot = TRUE)
@
<<echo = false>>=
RG(ex, M, C)
@
\begin{center}
\includegraphics[scale = 0.45]{fig3}
\end{center}
The summary graph is plotted with dashed undirected edges instead of bi-directed edges:
<<eval = false>>=  % 12
plotGraph(SG(ex,M,C), dashed = TRUE)
@
<<echo = false>>=
SG(ex, M, C)
@
\begin{center}
\includegraphics[scale = 0.45]{sg}
\end{center}
The induced ancestral graph is obtained from the DAG defined as a vector.
<<eval = false>>= % 13
AG(exvec,M,C,showmat=FALSE, plot=TRUE)
@
\begin{center}
\includegraphics[scale = 0.45]{fig4}
\end{center}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Verifying  $m$-separation }
To globally verifying  whether an independence statement of form $A\ci B \ci C$ is implied by a mixed graph we use a separation criterion called \emph{$m$-separation}. This has been defined in \citet{sad112} for the general class of loopless mixed  graphs and is the same as the $m$-separation criterion defined in \citet{ric02} for ancestral graphs. It is also a generalisation of the $d$-separation criterion for DAGs \citep{pea88}. This is
a graphical criterion that looks if the graph contains special paths connecting two sets $A$ and $B$
and involving a third set $C$ of the nodes. These special paths are said to be active or $m$-connecting.
For example, a directed path from a node in $A$ to  a node in $B$ that does not contain any node of $C$
is $m$ connecting $A$ and $B$. However, if such a path intercepts a node in $C$ then $A$ and $B$ are said to be
$m$-separated given $C$. However, this behaviour can change if the path connecting $A$ and $B$ contains
a collision node or a \emph{collider} for short, that is a node $c$
where the edges meet head-to-head, like e.g., $\fra c\fla$ or $\fra c \arc $.

In general, a path  is said to be \emph{$m$-connecting} given $C$ if all its
collider nodes are in $C$ or in the set of ancestors of $C$,  and all its non-collider nodes are outside $C$.
For two disjoint subsets of the node set $A$ and $B$, we say that $C$ $m$-separates $A$ and $B$
if there is no $m$-connecting path between $A$ and $B$ given $C$. This separation criterion
has been implemented in R and the related function is provided in \pkg{ggm}.
\paragraph{Examples.} The fundamental way to verify
if a graphical model implies an independence $A \cip B |C$ is to test whether
the disjoint subsets $A$ and $B$ of the node set are $m$-separated given a third set $C$, and
this may be done with the function \code{msep}. Note that there is still a function \code{dSep} in \pkg{ggm}
for $d$-separation, although it is superseded by \code{msep}.

The function has 4 arguments, where the first is the graph \code{a},
in one of the forms discussed before, and the other three are the disjoint sets
\code{A}, \code{B} and \code {C}.
For example, consider the DAG of Figure~\ref{dags}(a):
<<>>= % 18.2
a <- DAG(Y ~ U + Z, X ~ U + W, Z ~ W)
@


Then, we see that $Y$ and  $W$ are $m$-separated given $Z$:
<<>>= %18.3
msep(a,"Y","W","Z")
@
and the same statement holds for the induced
ancestral graph after marginalization over \code{U}:
<<>>= %18.4
b <- AG(a, M = "U")
msep(b,"Y","W","Z")
@
This was expected because the induced ancestral graph
respects all the $m$-separation statements present in the DAG, and
not involving the variable \code{U}.

As a more complex example, consider  the following mixed graph,
<<>>= % 19
a <- makeMG(dg= DG(W ~ Z, Z ~ Y + X),
            bg= UG(~ Y*Z))
@
<<eval = false>>= % 20
plotGraph(a)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setkeys{Gin}{width=0.3\textwidth}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\includegraphics[scale = 0.45]{fig5}
\end{center}
Then, the two following statements verify whether \code{X} is $m$-separated from \code{Y}
given \code{X}, and whether \code{X} is $m$-separated from \code{Y} (given the empty set):
<<>>= %21
msep(a,"X","Y","Z")
msep(a,"X","Y")
@
In this example $m$-separation is equivalent to its special case, the $d$-separation criterion
for DAGs.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Verifying maximality}
While for many subclasses of graphs a missing edge corresponds to some independence statement, for the more complex classes of mixed graphs this is not necessarily true. A graph where each of its missing edges is related to an independence statement is called a \emph{maximal graph}. For a more detailed discussion on maximality of graphs and graph-theoretical conditions for maximal graphs, see \citet{ric02} and \citet{sad11}.  \citet{sad11} also gave an algorithm for generating maximal ribbonless graphs that induce the same independence structure to an input non-maximal ribbonless graph. This algorithm has been
implemented here in \pkg{ggm} as illustrated below.

\paragraph{Function for generating maximal graphs.} Given a non-maximal graph, we can obtain the adjacency matrix of a
maximal graph that induces the same independence statements with the  function \code{Max}.
This function uses the algorithm by  \citet{sad113}, which is an extension of the implicit algorithm
presented in \citet{ric02}. The related functions
\code{MAG}, \code{MSG}, and \code{MRG}, are just handy wrappers to obtain maximal AGs, SGs and AGs, respectively.
For example,
<<>>= % 14
H <- matrix(c(0  ,100,  1,  0,
              100,0  ,100,  0,
              0  ,100,  0,100,
              0,  1  ,100,  0),4,4)
@
<<eval=FALSE>>= %15
plotGraph(H)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setkeys{Gin}{width=0.25\textwidth}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\includegraphics[scale = 0.45]{fig55}
\end{center}
is a mixed non-maximal graph, with the missing edge between nodes 1 and 4 that
is not associated with any independence statement.
Its associated maximal graph is obtained by
<<>>= %16
Max(H)
@
<<eval=FALSE>>= %17
plotGraph(Max(H))
@
\begin{center}
\includegraphics[scale = 0.45]{fig56}
\end{center}
As the graph \code{H} is an ancestral graph (as may be verified by the function
\code{isAG}), we obtain the same result with
<<>>= %18
MAG(H)
@




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Verifying Markov equivalence}
Two graphical models are said to be Markov equivalent when their associated graphs, although nonidentical, imply the same independence structure, that is the same set of  independence statements.   Thus  two Markov equivalent  models  cannot be discriminated on the basis of statistical tests of independence, even for arbitrary large samples.
For instance, the two directed acyclic graphs models $X \fla U \fra Y$ and
 $X \fla U \fla Y$  both  imply the same single independence statement between $X$ and $Y$ conditionally on $U$.

Sometimes, we can check whether graphs of different types are Markov equivalent.
For instance the DAG $X \fra U \fla Y$ is Markov equivalent to the bi-directed graph
$X \arc U \arc Z$.

Markov equivalent models may be useful in applications because (a) the may suggest alternative interpretations of a given well-fitting model or (b) on the basis of the
equivalence one can choose a simpler fitting algorithm. For instance, the
previous bi-directed graph model may be fitted, using the Markov equivalent DAG, in terms of a sequence of univariate regressions.

In the literature several problems related to Markov equivalences have been discussed. These include (a) verifying the Markov equivalence of given graphs, (b) presenting conditions under which a graph of a specific type can be Markov equivalent to a graph of another type, and (c) providing algorithms for generating Markov equivalent graphs of a certain type from a given graph.



\paragraph{Functions for testing Markov equivalences.} The function
\code{MarkEqRcg} tests whether two regression chain graphs
are Markov equivalent. This function simply
finds the skeleton and all unshielded collider V-configurations in both graphs
and tests whether they are identical, see \citet{wers11}. The arguments of this function are the two
graphs \code{a} and \code{b} in one of the allowed forms. For example,

<<>>=
H1 = makeMG(dg = DAG(`4` ~ `1`, `5` ~ `3`),
     bg = UG(~ `1`*`2` + `2`*`3` + `4`*`5`))
H2 = makeMG(dg = DAG(`4` ~ `1`, `5`~ `3`,
                     `2` ~ `1` + `3`),
            bg = UG(~ `4`*`5`))
H3 = DAG(`4`~`1`, `5`~`3` + `4`,`2`~`1` + `3`)
@
<<eval = false>>=
plotGraph(H1); plotGraph(H2); plotGraph(H3)
@
\begin{center}
\includegraphics[scale = 0.45]{rcg1}
\includegraphics[scale = 0.45]{rcg2}
\includegraphics[scale = 0.45]{rcg3}
\end{center}
We can now verify Markov equivalence as follows
<<>>=
MarkEqRcg(H1,H2)
MarkEqRcg(H1,H3)
MarkEqRcg(H2,H3)
@
To test Markov equivalence for maximal ancestral graphs the algorithm is computationally much more demanding (see \citet{ali04}) and, for this purpose, the
function \code{MarkEqMag} has been provided. Of course, one can use this function for Markov equivalence of regression chain graphs (which are a subclass of MAGs).  For example,
<<>>=
A1<- makeMG(dg = DG(`4`~`2`),
     bg =  UG(~ `1`*`2` + `2`*`3` + `3`*`4`))
A2<- makeMG(dg = DG(`4`~`2`, `2`~`1`),
     bg = UG(~ `2`*`3` + `3`*`4`))
A3<- makeMG(dg = DG(`4`~`2`, `2`~`1`, `3`~`1`),
     bg = UG(~ `3`*`4`))
@
<<eval = false>>=
plotGraph(A1); plotGraph(A2); plotGraph(A3)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setkeys{Gin}{width=0.2\textwidth}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\includegraphics[scale = 0.45]{ag1}\quad
\includegraphics[scale = 0.45]{ag2}\quad
\includegraphics[scale = 0.45]{ag3}
\end{center}
<<>>=
MarkEqMag(H1,H2)
MarkEqMag(H1,H3)
MarkEqMag(H2,H3)
@
\paragraph{Functions for generating Markov equivalent graphs of a specific type.}
Sometimes is important to verify if a given graph is capable of being Markov equivalent to a graph of a specific class of graphs (such as DAGs, undirected
graphs, or bidirected graphs), and if so, to obtain as a result such a graph.
The functions \code{RepMarDAG}, \code{RepMarUG}, and \code{RepMarBG} do this for DAGs, undirected graphs, or bidirected graphs, respectively. For associated conditions and algorithms, see \citet{sad113}. For example, given the following graph
<<>>=
H <- matrix(c( 0,10,  0,  0,
              10, 0,  0,  0,
               0, 1,  0,100,
               0, 0,100,  0),4,4)
@
<<eval=FALSE>>=
plotGraph(H)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setkeys{Gin}{width=0.5\textwidth}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\includegraphics[scale = 0.45]{fig57}
\end{center}
we can see that it is Markov equivalent to a DAG, by
<<>>=
RepMarDAG(H)
@
<<eval=FALSE>>=
plotGraph(RepMarDAG(H))
@
\begin{center}
\includegraphics[scale = 0.45]{fig58}
\end{center}
On the other hand it is not Markov equivalent to an undirected graph
or to a bidirected graph.
<<>>=
RepMarUG(H)
RepMarBG(H)
@
%This has consequence on the possibility of using
%standard fitting algorithms for DAG models to estimate
%the parameters of a model for the first graph.
\section{Acknowledgments}
The authors are grateful to Steffen Lauritzen for helpful suggestions on codes and comments on an earlier version of the paper and Nanny Wermuth and the referees for their insightful comments.
\bibliography{bib}



\end{document}